Histogram of Gradient Orientations of Signal Plots applied to P300 Detection

### 2017-12-20 response.letter.tex

Another issue with the manuscript is its mathematical vagueness. For example, the variable T is used as a vector or two-tuple but its notation is different from the notation the authors use for vectors. Moreover, there is something wrong in equation (4) and something inconsistent between equations (7) and (8). Also, the authors go back and forth the digital space and the continuous space and although there is an equivalence, the manuscript could benefit from a clearer explanation.

Another issue that I find worrisome is the design of experiments. First of all, it seems that the number of subjects is rather low, it would be interesting to know why the authors decided to use such a subset. Also, the authors speculate that their results are not that good because they did not take into account the latency and magnitude of the signals, there is previous evidence in the literature that the P300 suffers from these, and other, problems and, thus, one wonders why the authors did not anticipate such problems.

A couple of finale observations. First, it seems the authors assume that the readers have some previous deep knowledge of some topics such as that some channels should carry more or better information without a previous reference. Another issue has to do with the way the manuscript is written, sometimes is very difficult to follow the ideas the authors want to convey, an example can be found in the paragraph containing equation (5), the manuscript surely benefit from a careful review before submitting.

\subsection*{\ovalbox{General Comments}}
I have read the manuscript "EEG characterization and classification based on Histogram of Gradients of Signal Plots" with great interest since its title promised a solution to an important problem. In spite of presenting an interesting methodology and some interesting issues related to EEG characterization, I consider the manuscript is not ready for publication. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your comments and observations. We have modified several points according to your recommendations, we hope the article is now better suited for publication. 
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

2.1) the title is misleading since a first, and naive, read suggest being able to classify and characterize any EEG with their different components (i.e., P200, P300, P600, N100, N200, N400 to name a few). Clearly, the title does not match the content since the authors focus on the P300

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation. The title of the article was changed accordingly to your suggestion.  
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2.2) there is evidence of the benefits of working with these signals P300, but, is the proposed method extensible to other signals?, this is not clear from the manuscript.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you very much for your comment.  We published in ~\citep{Ramele2016} the application of the same method to identify rhythmic EEG events like Visual Occipital Alpha Waves and to classify Motor Imagery.  We are also working on unpublished material where we are analyzing the same approach for the detection of K-Complexes and classification of  SSVEP patterns but the results are in progress.  We added this information on the article (Introduction and Conclusion, section 3).  We hope this new information enriches the article and clarifies this point.

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
\subsection*{\ovalbox{Third Paragraph}}

3.1) Another issue with the manuscript is its mathematical vagueness. For example, the variable T is used as a vector or two-tuple but its notation is different from the notation the authors use for vectors

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  We worked hard to improve all the mathematical description of the method. The issue with variable T was fixed and we improved on the font styling to differentiate vectors and scalars.  Section 1.1 was modified as well as section 1.2.3.  The mathematical terminology were corrected all along the article. We hope we have full addressed this point. }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

3.2) Moreover, there is something wrong in equation (4)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We appreciate a lot your truthful comment.  Equation (4) is very important for us because it is the essence of the method.  %We based that equation on the description of how the histogram of oriented gradients is calculated published in Vedaldi and Szeliski, Computer Vision and Algorithms.  
We reviewed the equation, corrected their terms and we hope it is now more clear.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

3.3) and something inconsistent between equations (7) and (8)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We have revised the manuscript and modified equations (7) and (8) (now (8) and (9)) which are located on Section 1.2.3 (Parameters).  We added more information on the text and specified subscript indexes to avoid the inconsistency.
We thank you for noticing this.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3.4) Also, the authors go back and forth the digital space and the continuous space and although there is an equivalence, the manuscript could benefit from a clearer explanation.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  We modified all equations to emphasize the point that they are calculated in digital space, particularly on equation (4). We also added that image gradients are also calculated using finite differences, and added information about the support of each equation to explain that they are actually calculated in digital space.
With this, we hope we have full addressed your point.
 }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

\subsection*{\ovalbox{Fourth Paragraph}}

4.1) Another issue that I find worrisome is the design of experiments. It seems that the number of subjects is rather low, it would be interesting to know why the authors decided to use such a subset.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your comment. The original dataset was published as the 008-2014 dataset at the BNCI-Horizon website, which is very important in BCI Research.  As far as we are aware of, experiments on this dataset of ALS patients has not been published by any independent group.
In order to address this issue, we included a new own dataset with the same experimental conditions that we generated ourselves, on healthy subjects.  We published this dataset on the kaggle platform for replication.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

4.3) Also, the authors speculate that their results are not that good because they did not take into account the latency and magnitude of the signals, there is previous evidence in the literature that the P300 suffers from these, and other, problems and, thus, one wonders why the authors did not anticipate such problems.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your comments.  We added more detailed information that address this issue. First in Section 1.1.4  we included information about the signal averaging procedure.  In Section 1.2.3 we added the parameters that were used.  More importantly, in Section 2 we added information and references about the tests that we did to address latency and amplitude issues.  We included in Section 3 information about future and current works where we are also dealing with these problems.
We hope we full addressed your remarks.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

\subsection*{\ovalbox{Fifth Paragraph}}

5.1) it seems the authors assume that the readers have some previous deep knowledge of some topics such as that some channels should carry more or better information without a previous reference.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your suggestion. The Results section was modified to clarify the expectation about the EEG channels where a visual P300 oddball paradigm generates stronger signals.  We also added references.
We hope we  have addressed your point with this.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

5.2)  issue has to do with the way the manuscript is written

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you a lot for this important comment.  We re-read all the manuscript and asked  colleagues to proofread the article.  They gave us interesting suggestions that helped us to improve different sections.  We would be very pleased if you find it now easier to read it.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

5.3) sometimes is very difficult to follow the ideas the authors want to convey, an example can be found in the paragraph containing equation (5)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you very much for this information.  We reviewed completely that equation (it is now equation (6) and equation (7)). We included a step-by-step procedure that describes the proposed algorithm and we added a schematic chart in Figure 3.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

### 2017-12-20 article.tex
Agregamos todos los cambios de las conclusiones, correcciones generales sobre todo el texto, el cambio para hacer digital las señales, otros cambios en los gráficos, cambios en los textos y otros.

%=================================================================
\documentclass[entropy,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi} 
\usepackage{subfigure}

@@

This paper reports a method to, (1) classify P300 signals based on the identification of their structure in the shape domain using histograms of gradient orientations extracted from the image of signal plots, and (2) describe the way in which this classification can be used to implement an offline P300-based BCI Speller application using two public datasets. Its validity is verified by processing offline data for ALS patients and for healthy subjects.

This article unfolds as follows: Section~\ref{Feature} is dedicated to explain the Feature Extraction method based on Histogram of Gradient Orientations of the Signal Plot: Section~\ref{Pipeline} shows the preprocessing pipeline,  Section~\ref{Plot}  describes the image generation of the signal plot, Section~\ref{SIFT}  presents the feature extraction procedure while  Section~\ref{Classification}  introduces the whole  Speller Matrix Letter Identification procedure including the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008}. In Section~\ref{Protocol}, the experimental protocol is expounded. Section~\ref{Results} shows the results of applying the proposal technique and a discussion is presented. 
In the final Section~\ref{conclusion}  we expose our remarks, conclusions and future work.
%This kind of EEG events are described in~\citep{Hartman2005} and~\citep{Ben-Simon2013}.

@@

\begin{figure}[H]
\centering
\subfigure[Row flash.]{\includegraphics[width=.29\linewidth]{openvibep300matrix.png}}
\subfigure[Column flash.]{\includegraphics[width=.29\linewidth]{p300matrixcoloriginal.png}}
\caption{Example of the $6 \times 6$ Speller Matrix used in the study.  Rows and columns flash intermittently in random permutations.}
\label{fig:p300matrix}
\end{figure}

@@

In this section, the signal preprocessing, the method for generating images from signal plots, the feature extraction procedure and the speller matrix identification are described.

@@

\begin{itemize}
\item \textbf{Signal Enhancement}: The preprocessing stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. The processing pipeline starts by applying a notch filter to the raw digital signal, a 
$4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \citep{Krusienski2006}.
\item \textbf{Artifact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the $6$ rows plus the $6$ columns, a basic artefact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ stimuli,  a window of $t_{max} = 1$ second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments should contain the P300 ERP signature time-locked to the flashing stimulus, one for the row, and one for the column.
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out~\citep{Liang2008}. 
\end{itemize}

@@

\begin{equation}
\tilde{x}(n,c) = \left \lfloor{ \gamma \cdot \frac{( x(n,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor, n \in \{ 1, \dots, n_{max}\},c \in \{1,2,\dots,Ch\}
\label{eq:standarizedaverages}
\end{equation}

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related with the image scale. In addition, $ x(n,c) $ is the point-to-point averaged EEG matrix for the sample point $n$ and for channel $c$, $n_{max}=F_s.t_{max}$ and $F_s$ is the sampling frequency. The parameter  $Ch$ is the number of available EEG channels. Lastly, $$\bar{x}(c) =\frac{1}{n_{max}}\sum_{n=1}^{n_{max}}x(n,c)$$ and $$ \sigma(c) = (\frac{1}{n_{max}-1}\sum_{n=1}^{n_{max}}(x(n,c)-\bar{x}(c))^2 )^{\frac{1}{2}}$$ are the mean and standard deviation of $x(n,c), n \in 1,\dots,n_{max}$, for each channel $c$.

@@

\noindent where $ (z_1,z_2) \in \mathbb{N} \times \mathbb{N}$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image,  $n \in \{1, \dots, n_{max}\}$ and $c \in \{1,2,\dots,Ch\}$.  The function $z(c)$, $c \in \{1,2,\dots,Ch\}$ is the location on the image where the signal's zero value has to be located in order to fit the entire signal within the image:

@@

\begin{equation}
 h(\theta,i,j) = 3s \sum_{\mathbf{x} \in B_{i,j}} w_\mathrm{ang}(\angle J(\mathbf{x}) - \theta)\, w_{ij}\left(\frac{\mathbf{x} - \mathbf{kp}}{3 s}\right)\, |J(\mathbf{x})|
\label{eq:histogram}
\end{equation}

\noindent  where $\mathbf{x}$ is a pixel from  the $i,j$-block $B_{i,j}$, $ i,j \in \{1,2,3,4\} $, $ \theta \in \{0, 45, 90, 135, 180, 225, 270, 315\} $,  $ |J(\mathbf{x})| $ is the norm of the gradient vector in the pixel $\mathbf{x}$ and it is computed using finite differences, $\angle J(\mathbf{x}) $ is the angle of the gradient vector,  $\theta$ is the angle bin,   and $ w_\mathrm{ang}(\cdot) $  and $ w_{ij}(\cdot) $ are linear interpolation functions used by Lowe and Vedaldi et al. in~\citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each block when $s = 1$.  As the patch has  $16$ blocks and  $8$ bin angles are considered, a descriptor of $128$ dimension is obtained. It can be observed that in each step, the histogram is computed by multiplying by $ |J(\mathbf{x})| $, so the method considers both, the magnitude and the orientation of the gradient vector.

@@

\begin{figure}[H]
\centering
\subfigure[Plot of the signal, a keypoint and the corresponding patch.]{\includegraphics[width=.49\linewidth]{signalpatchkeypoint.png}\label{patch}}
\subfigure[Orientations on two blocks of the patch.]{\includegraphics[width=.49\linewidth]{histogramsbines.png}\label{histogrambines}}
%\includegraphics[width=9cm]{gradientszoom.png}
\caption{ Example of a patch and a scheme of the orientation's histogram computation.}
\label{fig:sampledescriptor}
\end{figure}

@@

The aim is to identify the selected letter from the matrix. Previously, during training phase, two descriptors are extracted from averaged signal segments which correspond to the letter where the user was supposed to be focusing onto.  These descriptors are the P300 templates which are grouped in a template set called $ T $.  Segments corresponding to rows are labeled 1-6, whereas those corresponding to columns are labeled 7-12.  The process has the following steps:

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the single trial segments $S_1(n,c),\dots,S_{k_a}(n,c)$, where the variables $n \in \{1, \dots, n_{max}\}$ and $c \in \{1,2,\dots,Ch\}$ correspond to time and channel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item \label{paso3} Compute the Ensemble Average by
\begin{equation}
x(n,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(n,c),n \in \{1, \dots, n_{max}\}, c \in \{1,\dots,Ch\}
\label{averaging}
\end{equation}  
for each row and column.

@@

where $NN_T(d^l_u), \;l\in\{row,{col}\}$ is the set of the $k$ nearest neighbors to $d^l_u, \;l\in\{row,{col}\}$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l\in\{row,{col}\}$.

@@

\end{enumerate}
By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{row} $ and column $ \hat{col} $. 
Figure~\ref{fig:classification} shows a scheme of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments $S_i$ are averaged for the 6 rows and 6 columns. From the averaged signal, the image of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure}

@@

Participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and all participants agree and sign a written informed consent.  All healthy subjects have normal or corrected-to normal vision and no history of neurological disorders. The experiment is performed with 8 subjects, 6 males, 2 females, 6 right-handed, 2 left-handed, average age 28.25 years, standard deviation  9.58 years, range 20-50 years.

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole transient event.  The size of the patch is $S_p \times S_p$ pixels. The vertical size consists of $4$ blocks of size $3 s_y$ pixels which is long enough as to contain the signal $\Delta  \mu V $, the peak-to-peak amplitude of the transient event. The horizontal size on the other hand, includes $4$ blocks  of $3 s_x$ and covers the entire duration in seconds of the transient signal event, $ \lambda $.   }

@@

Lastly, the number of channels $Ch$ is equal to $8$ for both datasets, and the number of intensification sequences $k_a$ is statically assigned to $10$.  The parameter $k$ used in the Near Neighbor $NN_T(d^l_u), \;l\in\{row,{col}\}$ is set to $k=7$, following the suggestion of the article~\citep{Boiman2008}.  In addition, the norm used on  Equations \ref{eq:multiclassificationrow} and \ref{eq:multiclassificationcol} was the cosine norm, and descriptors were normalized to $ \left[ -1, 1 \right] $.

@@

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each subject of the public dataset 008-2014. Twenty letters, four words of five letters each, were used for testing. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &   $35\%$  \\
2     &     Fz   &   $85\%$  \\
3     &     Cz   &   $25\%$  \\
4     &     PO8 &   $55\%$  \\
5     &     PO7 &   $40\%$ \\
6     &     PO7 &   $60\%$  \\
7     &     PO8 &   $80\%$  \\
8     &     PO7 &   $95\%$ \\

@@

In Table~\ref{tab:resultsowndataset} results obtained for 8 healthy subjects are shown.   These results were obtained by the present method as described in previous sections.

@@

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each healthy subject. The method is applied on a channel by channel basis, and the best performing channel is shown. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{cccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Oz   &     $40\%$  \\
2     &     PO7   &     $30\%$ \\
3     &     P4   &     $40\%$ \\
4     &     P4   &     $45\%$ \\
5     &     P4   &      $60\%$ \\
6     &     Pz   &      $50\%$ \\
7     &     PO7   &      $70\%$ \\
8     &     P4   &      $50\%$ \\

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{performance2.eps}
\caption{Performance curves for the eight subjects included in the dataset of ALS patients.  Three out of eight subjects achieved the necessary performance to implement a valid P300 speller.}
\label{fig:performance}
\end{figure}

As can be seen in the figure, as the number of intensification sequences tend to 1, which corresponds to single trial letter identification, the performance is reduced. As mentioned before, the SNR of the single trial P300 is very low. The ensemble average is carried out in order to improve the SNR but other problems arise, for example inter-trial variability due to latencies shifts that may produce an unstable P300 signature. To solve this problem, we tested an approach to assess if the morphological shape of the P300 can be stabilized by applying different shifts and we verified that there is a better performance when a correct resynchronization is applied, but results are still in process.  We also applied Dynamic Time Warping (DTW)~\citep{Casarotto2005} but we were unable to find a substantial improvement.     On the other hand amplitude, width and latency are generally affected by habituation, fatigue or level of attention and may lead to null signals\citep{Ouyang2017}. This is another source of instability of the P300 signature component that may need to be addressed.

>>>>

%=================================================================
\documentclass[entropy,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi} 
\usepackage{subfigure} 
\usepackage{siunitx}

@@

This paper reports a method to, (1) classify P300 signals based on the identification of their structure in the shape domain using histograms of gradient orientations extracted from the image of signal plots, and (2) describe the way in which this classification can be used to implement an offline P300-based BCI Speller application using two datasets. Its validity is verified by processing offline data for ALS patients and for healthy subjects.

This article unfolds as follows: Section~\ref{Feature} is dedicated to explain the Feature Extraction method based on Histogram of Gradient Orientations of the Signal Plot: Section~\ref{Pipeline} shows the preprocessing pipeline,  Section~\ref{Plot}  describes the image generation of the signal plot, Section~\ref{SIFT}  presents the feature extraction procedure while  Section~\ref{Classification}  introduces the whole  Speller Matrix Letter Identification procedure including the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008}. In Section~\ref{Protocol}, the experimental protocol is expounded. Section~\ref{Results} shows the results of applying the proposal technique and a discussion is presented as well.
In the final Section~\ref{conclusion}  we expose our remarks, conclusions and future work.
%This kind of EEG events are described in~\citep{Hartman2005} and~\citep{Ben-Simon2013}.

@@

\begin{figure}[htb]
\centering
\subfigure[Row flash.]{\includegraphics[width=.29\linewidth]{openvibep300matrix.png}}
\subfigure[Column flash.]{\includegraphics[width=.29\linewidth]{p300matrixcoloriginal.png}}
\caption{Example of the $6 \times 6$ Speller Matrix used in the study.  Rows and columns flash intermittently in random permutations.}
\label{fig:p300matrix}
\end{figure}

@@

In this section, the signal preprocessing, the method for generating images from signal plots, the feature extraction procedure and the Speller Matrix identification are described.

@@

\begin{itemize}
\item \textbf{Signal Enhancement}: The preprocessing stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. The processing pipeline starts by applying a notch filter to the raw digital signal, a 
$4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \citep{Krusienski2006}.
\item \textbf{Artifact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every complete sequence of intensification of $6$ rows and $6$ columns, a basic artifact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ stimuli,  a window of $t_{max} = 1$ second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments should contain the P300 ERP signature time-locked to the flashing stimulus, one for the row, and one for the column.
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out~\citep{Liang2008}. 
\end{itemize}

@@

\begin{equation}
\tilde{x}(n,c) = \left \lfloor{ \gamma \cdot \frac{( x(n,c) - \bar{x}(c)  )}{ \hat{\sigma}(c) } }\right \rfloor, \; n \in \{ 1, \dots, n_{max}\},\; c \in \{1,2,\dots,Ch\}
\label{eq:standarizedaverages}
\end{equation}

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related to the image scale. In addition, $ x(n,c) $ is the point-to-point averaged EEG matrix for the sample point $n$ and for channel $c$, $n_{max}=F_s.t_{max}$ and $F_s$ is the sampling frequency. The parameter  $Ch$ is the number of available EEG channels. Lastly, $$\bar{x}(c) =\frac{1}{n_{max}}\sum_{n=1}^{n_{max}}x(n,c)$$ and $$ \hat{\sigma}(c) = (\frac{1}{n_{max}-1}\sum_{n=1}^{n_{max}}(x(n,c)-\bar{x}(c))^2 )^{\frac{1}{2}}$$ are the mean and estimated standard deviation of $x(n,c), n \in 1,\dots,n_{max}$, for each channel $c$.

@@

%$\max\limits_{1\leq j\leq n}$

\noindent where $ (z_1,z_2) \in \mathbb{N} \times \mathbb{N}$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image,  $n \in \{1, \dots, n_{max}\}$ and $c \in \{1,2,\dots,Ch\}$.  The values $z(c)$, $c \in \{1,2,\dots,Ch\}$ are the location on the image where the signal's zero value has to be located in order to fit the entire signal within the image for each c:

@@

\begin{equation}
 h(\theta,i,j) = 3s \sum_{\mathbf{p} \in B_{i,j}} w_\mathrm{ang}(\angle J(\mathbf{p}) - \theta)\, w_{ij}\left(\frac{\mathbf{p} - \mathbf{kp}}{3 s}\right)\, |J(\mathbf{p})|
\label{eq:histogram}
\end{equation}

\noindent  where $\mathbf{p}$ is a pixel from  the $i,j$-block $B_{i,j}$, $ i,j \in \{1,2,3,4\} $,  $\theta$ is the angle bin with $ \theta \in \{0, 45, 90, 135, 180, 225, 270, 315\} $,  $ |J(\mathbf{p})| $ is the norm of the gradient vector in the pixel $\mathbf{p}$ and it is computed using finite differences, $\angle J(\mathbf{p}) $ is the angle of the gradient vector, and $ w_\mathrm{ang}(\cdot) $  and $ w_{ij}(\cdot) $ are linear interpolation functions used by Lowe and Vedaldi et al. in~\citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each block when $s = 1$.  As the patch has  $16$ blocks and  $8$ bin angles are considered, a descriptor of $128$ dimension is obtained. It can be observed that in each step, the histogram is computed by multiplying by $ |J(\mathbf{p})| $, so the method considers both, the magnitude and the orientation of the gradient vector.

@@

\begin{figure}[htb]
\centering
\subfigure[Plot of the signal, a keypoint and the corresponding patch.]{\includegraphics[width=.49\linewidth]{signalpatchkeypoint.png}\label{patch}}
\subfigure[Orientations on two blocks of the patch.]{\includegraphics[width=.49\linewidth]{histogramsbines.png}\label{histogrambines}}
%\includegraphics[width=9cm]{gradientszoom.png}
\caption{ Example of a patch and a scheme of the orientation's histogram computation.}
\label{fig:sampledescriptor}
\end{figure}

@@

The aim is to identify the selected letter from the matrix. Previously, during the training phase, two descriptors are extracted from averaged signal segments which correspond to the letter where the user was supposed to be focusing onto.  These descriptors are the P300 templates which are grouped in a template set called $ T $.  This set is constructed using the steps 1-6 of the following algorithm.  Segments corresponding to rows are labeled 1-6, whereas those corresponding to columns are labeled 7-12.  The whole process has the following steps:

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the single trial segments $S_1(n,c),\dots,S_{k_a}(n,c)$, where the variables $n \in \{1, \dots, n_{max}\}$ and $c \in \{1,2,\dots,Ch\}$ correspond to sample points and channel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item \label{paso3} Compute the Ensemble Average by
\begin{equation}
x(n,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(n,c),n \in \{1, \dots, n_{max}\}, c \in \{1,\dots,Ch\}
\label{averaging}
\end{equation}  
for each row and column.

@@

where $NN_T(d^l_u), \;l\in\{row,{col}\}$ is the set of the $k$ nearest neighbors to $d^l_u, \;l\in\{row,{col}\}$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l\in\{row,{col}\}$.  This set is obtained according to the k-NNBN  algorithm~\citep{Boiman2008}.

@@

\end{enumerate}
By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{row} $ and column $ \hat{col} $. 
Figure~\ref{fig:classification} shows a scheme of this process.

\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments $S_i$ are averaged for the 6 rows and 6 columns. From the averaged signal, the image of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure}

@@

Participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and all participants agree and sign a written informed consent.  All healthy subjects have normal or corrected-to normal vision and no history of neurological disorders. The experiment is performed with 8 subjects, 6 males, 2 females, 6 right-handed, 2 left-handed, average age 29.00 years, standard deviation  11.56 years, range 20-56 years.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole transient event.  The size of the patch is $S_p \times S_p$ pixels. The vertical size consists of $4$ blocks of size $3 s_y$ pixels which is long enough as to contain the signal $\Delta  \mu V $, the peak-to-peak amplitude of the transient event. The horizontal size includes $4$ blocks  of $3 s_x$ and covers the entire duration in seconds of the transient signal event, $ \lambda $.   }

@@

Lastly, the number of channels $Ch$ is equal to $8$ for both datasets, and the number of intensification sequences $k_a$ is statically assigned to $10$.  The parameter $k$ used in the Near Neighbor $NN_T(d^l_u), \;l\in\{row,{col}\}$ is set to $k=7$, following the suggestion of the article~\citep{Boiman2008}.  In addition, the norm used on  Equations \ref{eq:multiclassificationrow} and \ref{eq:multiclassificationcol} is the cosine norm, and descriptors are normalized to $ \left[ -1, 1 \right] $.

@@

\begin{table}[htb]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each subject of the public dataset 008-2014. The spelled words are \textit{GATTO}, \textit{MENTE}, \textit{VIOLA} and \textit{REBUS}. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &   $35\%$  \\
2     &     Fz   &   $85\%$  \\
3     &     Cz   &   $25\%$  \\
4     &     PO8 &   $55\%$  \\
5     &     PO7 &   $40\%$ \\
6     &     PO7 &   $60\%$  \\
7     &     PO8 &   $80\%$  \\
8     &     PO7 &   $95\%$ \\

@@

In Table~\ref{tab:resultsowndataset} results obtained for 8 healthy subjects are shown.

@@

\begin{table}[htb]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each healthy subject. The spelled words are \textit{MANSO},\textit{CINCO},\textit{JUEGO} and \textit{QUESO}. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{cccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Oz   &     $40\%$  \\
2     &     PO7   &     $30\%$ \\
3     &     P4   &     $40\%$ \\
4     &     P4   &     $45\%$ \\
5     &     P4   &      $60\%$ \\
6     &     Pz   &      $50\%$ \\
7     &     PO7   &      $70\%$ \\
8     &     P4   &      $50\%$ \\

@@

\begin{figure}[htb]
\centering
\includegraphics[width=10cm]{performance2.eps}
\caption{Performance curves for the eight subjects included in the dataset of ALS patients.  Three out of eight subjects achieved the necessary performance to implement a valid P300 speller.}
\label{fig:performance}
\end{figure}

As can be seen in the figure, as the number of intensification sequences tend to 1, which corresponds to single trial letter identification, the performance is reduced. As mentioned before, the SNR of the single trial P300 is very low. The ensemble average is carried out in order to improve the SNR but other problems arise, for example inter-trial variability due to latencies shifts that may produce an unstable P300 signature. To solve this problem, we tested an approach to assess if the morphological shape of the P300 can be stabilized by applying different shifts and we verified that there is a better performance when a correct single-trial alignment is applied, but results are still in process.  We also applied Dynamic Time Warping (DTW)~\citep{Casarotto2005} but we were unable to find a substantial improvement.     On the other hand amplitude, width and latency are generally affected by habituation, fatigue or level of attention and may lead to null signals\citep{Ouyang2017}. This is another source of instability of the P300 signature component that may need to be addressed.
